---
title: "data-day2-analyses"
author: "Jamil Bhanji and Vanessa Lobue"
date: "12/1/2022"
output: html_document
---

#### set up: set root directory and specify packages ("tidyverse")  
```{r setup, include=TRUE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
library(tidyverse)
#command to change data viewing preference to show more columns in the data viewer:
rstudioapi::writeRStudioPreference("data_viewer_max_columns", 500L)
```

*thanks to Darian for volunteering data!!!*  

#### Import data and data dictionary from tab-delimited file 
- and set variable types  
```{r import-data, include=TRUE}
# read data and data dictionary
dr_tib <- readr::read_delim("data/2022/Darian-Raizberg-DataSet.txt",
                              delim = "\t")
dr_dict <- readr::read_delim("data/2022/Darian-Raizberg-Data-Dictionary.txt", delim = "\t")
#recode Test_setting, Group, accuracy, choice variables as factors
dr_tib <- dr_tib %>% mutate(
    Test_setting = factor(Test_setting, labels = c("in-person","online")),
    Group = factor(Group, labels = c("high effort","low effort","no effort")),
    Part2_Phase1_acc = factor(Part2_Phase1_acc, labels = c("incorrect","correct")),
    Part2_Phase2_acc = factor(Part2_Phase2_acc, labels = c("incorrect","correct")),
    Part2_Phase3_acc = factor(Part2_Phase3_acc, labels = c("incorrect","correct")),
    Part2_Phase2_fdbk_choice = factor(Part2_Phase2_fdbk_choice, 
                                      labels = c("not chosen","chosen"))
  ) 
sapply(dr_tib,class) %>%  knitr::kable(caption = "variable names and types") %>% 
  kableExtra::kable_styling()
```


### Compute variables and visualize      
- Data consist of many trials per subject, is this what we want? (aggregate?)   

```{r aggregate-vis, results='hold', fig.show='hold'}
# first group by participants and store in new tibble
dr_aggbysub_tib <- dr_tib %>%
  drop_na(Part2_Phase2_acc,Part2_Phase3_acc,Part2_Phase2_fdbk_choice) %>% 
  group_by(Subj_ID,Test_setting,Group) %>% 
  summarise (
    numtrials = sum(Part2_Phase2_fdbk_choice=="chosen"|Part2_Phase2_fdbk_choice=="not chosen"),
    fdbk_choice_prop = sum(Part2_Phase2_fdbk_choice=="chosen")/numtrials,
    Part2_Phase2_acc_prop = sum(Part2_Phase2_acc=="correct")/numtrials,
    Part2_Phase3_acc_prop = sum(Part2_Phase3_acc=="correct")/numtrials,
    Part2_Phase2_conf_subjmean = mean(Part2_Phase2_conf)
    #also RT?
    #Part2_Phase2_RT_subjmean = mean(Part2_Phase2_RT)
  ) %>% ungroup()
# then print table of means, etc across participants, 
dr_aggbysub_tib %>% select(numtrials:last_col()) %>% 
  psych::describe() %>% knitr::kable(caption = "Whole sample descriptives", digits = 2) %>% 
  kableExtra::kable_classic(lightable_options = "hover")
# cell counts
dr_aggbysub_tib %>% count(Test_setting,Group)
# no points meet the exclusion criteria, but if they did, apply a filter and re-compute the descriptives

# data visualization for relevant vars
p1 <- dr_aggbysub_tib %>% #drop_na() %>%
  ggplot( aes(x=fdbk_choice_prop)) + 
    geom_histogram(position = "identity", binwidth = .1, alpha = .25) + 
    facet_wrap(~ Test_setting + Group) +
    theme_classic() + labs (title = "fdbk_choice_prop histogram")
p1
p2 <- dr_aggbysub_tib %>% #drop_na() %>%
  ggplot( aes(y=fdbk_choice_prop, x = Test_setting:Group)) + 
    geom_boxplot() + 
    theme_classic() +labs (title = "fdbk_choice_prop box plot")
p2
p3 <- dr_aggbysub_tib %>% #drop_na() %>%
  ggplot( aes(sample=fdbk_choice_prop)) + 
    geom_qq() + geom_qq_line() + 
    theme_classic() + labs (title = "fdbk_choice_prop Q-Q")
p3

```


### 2x2 Anova (between subjects)  

```{r two-by-two, results='hold', fig.show='hold'}
dr_2x2_afex <- dr_aggbysub_tib %>% 
  afex::aov_4(formula = fdbk_choice_prop ~ Test_setting*Group + (1|Subj_ID))
dr_2x2_afex
# or equivalently
dr_2x2_afex <- afex::aov_ez("Subj_ID", "fdbk_choice_prop",
               between = c("Test_setting","Group"), data = dr_aggbysub_tib)
dr_2x2_afex
afex::afex_plot(dr_2x2_afex)
# but notice that the aov() function gives different stats for the main effects, because it is based on type 1/sequential SS (rather than type 3) 
dr_2x2_aov <- aov(fdbk_choice_prop ~ Test_setting*Group, 
                  data = dr_aggbysub_tib)
anova(dr_2x2_aov)
#but the model and residuals are the same, so you can use the plot() function for model diagnostics
plot(dr_2x2_aov)

#with confidence covariate the model is Rank deficient (can't be estimated, usually not enough data per cell)
dr_2x2cov_afex <- afex::aov_ez("Subj_ID", "fdbk_choice_prop",
               between = c("Test_setting","Group"), 
               covariate = "Part2_Phase2_conf_subjmean", 
               data = dr_aggbysub_tib)
dr_2x2cov_afex
```


## Adding a within subjects factor  
- need to re-aggregate and include the addnl factor  
```{r aggregate-vis, results='hold', fig.show='hold'}
# first group by participants and store in new tibble
dr_aggbysubcost_tib <- dr_tib %>%
  drop_na(Part2_Phase2_acc,Part2_Phase3_acc,Part2_Phase2_fdbk_choice) %>% 
  group_by(Subj_ID,Test_setting,Group,Part2_Phase2_fdbk_cost) %>% 
  summarise (
    numtrials = sum(Part2_Phase2_fdbk_choice=="chosen"|Part2_Phase2_fdbk_choice=="not chosen"),
    fdbk_choice_prop = sum(Part2_Phase2_fdbk_choice=="chosen")/numtrials,
    Part2_Phase2_acc_prop = sum(Part2_Phase2_acc=="correct")/numtrials,
    Part2_Phase3_acc_prop = sum(Part2_Phase3_acc=="correct")/numtrials,
    Part2_Phase2_conf_subjmean = mean(Part2_Phase2_conf)
    #also RT?
    #Part2_Phase2_RT_subjmean = mean(Part2_Phase2_RT)
  ) %>% ungroup()


dr_2x2x4_afex <- afex::aov_ez("Subj_ID", "fdbk_choice_prop",
               between = c("Test_setting","Group"), 
               within = "Part2_Phase2_fdbk_cost", data = dr_aggbysubcost_tib)
dr_2x2x4_afex
afex::afex_plot(dr_2x2x4_afex)
```
```{r}
dr_aggbysub_tib %>% select(fdbk_choice_prop,Part2_Phase2_conf_subjmean) %>% correlation::correlation()

```

