---
title: "data-day2-analyses"
author: "Jamil Bhanji and Vanessa Lobue"
date: "12/1/2022"
output: html_document
---

#### set up: set root directory and specify packages ("tidyverse")  
```{r setup, include=TRUE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
library(tidyverse)
#command to change data viewing preference to show more columns in the data viewer:
rstudioapi::writeRStudioPreference("data_viewer_max_columns", 500L)
```

*thanks to Darian for volunteering data!!!*  

#### Import data and data dictionary from tab-delimited file  
- and set variable types  
```{r import-data, include=TRUE}
dr_tib <- readr::read_delim("data/2022/Darian-Raizberg-DataSet.txt",
                              delim = "\t")
dr_dict <- readr::read_delim("data/2022/Darian-Raizberg-Data-Dictionary.txt", delim = "\t")
dr_tib <- dr_tib %>% mutate(
    Test_setting = factor(Test_setting, labels = c("in-person","online")),
    Group = factor(Group, labels = c("high effort","low effort","no effort")),
    Part2_Phase2_acc_fac = factor(Part2_Phase2_acc, "incorrect","correct"),
    Part2_Phase3_acc_fac = factor(Part2_Phase3_acc, "incorrect","correct")
  )
sapply(dr_tib,class) %>%  knitr::kable(caption = "variable names and types") %>% 
  kableExtra::kable_styling()
```

#### Hypotheses/Research Questions and planned tests  

- *What is the hypothesis examined? What variables are used to test the hypothesis?*  
1. Participants that complete the more effortful task online will have decreases in their feedback-purchasing behaviors compared to those that complete the task in-person.      
2. Participants that get the no effort manipulation will not significantly differ in feedback-purchasing behavior as a function of task setting (no effort in-person vs no effort online).   
    - rephrase 1 and 2 as an interaction of effort by setting?  
    
3. Participant’s response confidence will correlate with their likelihood to purchase feedback.     - discuss differences btw by-participant approach and by-trial approach  

- *What statistical tests will you use? What assumptions do they make?*  
    - factorial ANOVA  
    - assume normality, homogeneity of residuals     

- *How will you treat extreme values? Missing data?*  
    - exclude participants 

##### Descriptives and distributions 

1.  Check cases and missing values, apply any exclusion criteria:

    -   Are missing values appearing appropriately? (scan the table after importing)\
    -   How many full cases?\
    -   Do missing data raise any concerns?  

2.  Mean, median, sd, min, max, cases for numeric variables of interest\

3.  Category counts for categoricals\

4.  Box plots, histograms for vars of interest (Q-Q plots if you want)

# before anything, check data types of variable, compute necessary variables and recode anything that needs to be recoded  
