---
title: "data-day2-analyses"
author: "Jamil Bhanji and Vanessa Lobue"
date: "12/1/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
library(ggplot2)
library(tidyverse)
```

## Omar's Data

```{r import-omar}
#command to change data viewing preference to show more columns:
#rstudioapi::writeRStudioPreference("data_viewer_max_columns", 1000L)

omar_tib <- readr::read_delim(delim="\t",file="data/omaralbastaki_200306_19599636_Omar Albastaki_Dataset-1.txt") %>% 
  janitor::clean_names()


```

#### Hypothesis/Research Question

-   what is the hypothesis examined?

    -   exposure to bad news about a brand decreases liking of the brand, but less so for brand that consumers are loyal to (interaction of loyalty and bad news on ratings)

-   look at the data - what columns are used to test the hypothesis?

    -   what is the DV?

    -   what are the IVs?

    -   any covariates?

-   do we need to reformat the data?

```{r format-columns}
# drop empty rows (no id)
omar_tib <- omar_tib %>% drop_na(id)

#1. create two columns: 1 to code loyal/random condition and 1 to code misaligned/aligned condition  "|" means "OR", "!" means "not"
omar_tib <- omar_tib %>% rowwise() %>% 
  mutate(
    loyal = if_else( 
      (!is.na(loyal_misaligned_rating)) | !(is.na(loyal_aligned_rating)), 
      1, 0),
    misaligned = if_else( 
      (!is.na(loyal_misaligned_rating)) | !(is.na(random_misaligned_rating)), 
      1, 0) 
  ) %>% ungroup()

#2. Create one DV column, called finalrating
omar_tib <- omar_tib %>% 
  mutate(
    final_rating = case_when(
      (loyal == 1) & (misaligned == 1) ~ loyal_misaligned_rating,
      (loyal == 1) & (misaligned == 0) ~ loyal_aligned_rating,
      (loyal == 0) & (misaligned == 1) ~ random_misaligned_rating,
      (loyal == 0) & (misaligned == 0) ~ random_aligned_rating
    )
  )
# 2.1 Create a column with the change score (final rating - initial rating)
omar_tib <- omar_tib %>% 
  mutate(
    rating_change = case_when(
      (loyal == 1) ~ final_rating - loyal_rating,
      (loyal == 0) ~ final_rating - random_rating
    )
  )

#now check the data
dplyr::glimpse(omar_tib)

```

##### Descriptives and examine distributions of relevant variables

1.  Check cases and missing values, apply any exclusion criteria:

    -   Are missing values appearing appropriately? (scan the table after importing)\
    -   How many full cases?\
    -   Are you concerned about missing data?

2.  Mean, median, sd, min, max, cases for numeric variables of interest\

3.  Category counts for categoricals\

4.  Box plots, histograms for vars of interest (Q-Q plots if you want)

```{r omar-vis, include=TRUE, fig.show='asis'}
# get descriptives for numerical vars
omar_tib %>% select(loyal_rating,random_rating,final_rating,rating_change) %>% 
  psych::describe() %>% 
  knitr::kable(caption = "numerical variable descriptives", digits = 2) %>% 
  kableExtra::kable_classic(lightable_options = "hover")


# data visualization for relevant vars
p1 <- omar_tib %>% 
  ggplot( aes(x=final_rating)) + 
    geom_histogram(position = "identity", binwidth = 1, 
                   alpha = .5) + 
    theme_classic() + labs (title = "final_rating histogram")
p1
p2 <- omar_tib %>% 
  ggplot( aes(x=rating_change)) + 
    geom_histogram(position = "identity", binwidth = 1, 
                   alpha = .5) + 
    theme_classic() + labs (title = "rating change histogram")
p2
p3 <- omar_tib %>% #drop_na() %>%
    ggplot( aes(y=final_rating)) + 
      geom_boxplot() + 
      theme_classic() +
      labs (title = "final_rating box plot")
p3
p4 <- omar_tib %>% #drop_na() %>%
    ggplot( aes(y=rating_change)) + 
      geom_boxplot() + 
      theme_classic() +
      labs (title = "final_rating box plot")
p4

omar_bygroup <- omar_tib %>% mutate(
  loyal = factor(loyal, levels = c(0,1), 
                 labels = c("random","loyal")),
  misaligned = factor(misaligned, levels = c(0,1), 
                      labels = c("aligned","misaligned")),
  ) %>% 
  group_by(loyal,misaligned) %>% 
  dplyr::summarise(
    mean_rating_change = mean(rating_change, na.rm=TRUE),
    ci.low = ggplot2::mean_cl_normal(rating_change)$ymin,
    ci.upp = ggplot2::mean_cl_normal(rating_change)$ymax,
  ) %>% ungroup()

p5 <- omar_bygroup %>% 
  ggplot( aes( x=misaligned, y=mean_rating_change, fill=loyal, ymin=ci.low, ymax=ci.upp) ) +
    geom_bar(stat = "identity", position=position_dodge(.9)) +
    geom_errorbar(width=.25, position=position_dodge(.9)) +
    theme_classic() + labs(title="rating change by loyalty and alignment", y = "rating change", x = "alignment", fill="brand loyalty")
p5

```

##### Statistical model

-   outcome: rating_change
-   independent vars (predictors): loyal, misaligned
-   model (formula): rating_change \~ loyal + misaligned + loyal:misaligned

```{r omar-anova}

#statistical model using the afex (analysis of factorial experiments) package
rating_change_afx <- afex::aov_4(
  rating_change ~ loyal + misaligned + loyal:misaligned + (1|id),
  data = omar_tib)
rating_change_afx

#simple effects
cat("\nsimple effects- effect of misalign at each level of loyal:\n")
emmeans::joint_tests(rating_change_afx, "loyal")

#also run it with lm() so we can look at the model residuals (not possible with the afex ouput type)
#why didn't we run it with lm() in the first place? because partitioning variance with anova() after lm() will partition variance sequentially (type 1 SS) but what we want is each main effect assigned variance considering the other main effect and interaction (type 3 SS), which is why we use afex::aov_4()
rating_change_lm <- lm(
  rating_change ~ loyal + misaligned + loyal:misaligned,
  data = omar_tib)

#model diagnostics - residual plots
plot(rating_change_lm)

```

##### export to a csv file:

```{r export-csv}
omar_tib %>% readr::write_csv("data/omar_data_recoded.csv")
```
