---
title: "data-day1-analyses"
author: "Jamil Bhanji and Vanessa Lobue"
date: "10/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
library(ggplot2)
library(tidyverse)
```

## Zion's Data

```{r import-zion}
#command to change data viewing preference to show more columns:
#rstudioapi::writeRStudioPreference("data_viewer_max_columns", 1000L)

zion_tib <- readr::read_delim(delim="\t",file="data/crichlowzion_52153_19605401_Zion-Crichlow-dataset.txt")


```

#### Hypothesis/Research Question 1

-   what is the hypothesis examined? what columns are used to test the hypothesis

    -   people have a stress response to their own violent behavior

    -   girls will have a larger stress response to their own violent behavior

    -   race?

    -   

##### Descriptives and examine distributions of relevant variables

1.  Check cases and missing values, apply any exclusion criteria:

    -   Are missing values appearing appropriately? (scan the table after importing)\
    -   How many full cases?\
    -   Are you concerned about missing data?

2.  Mean, median, sd, min, max, cases for numeric variables of interest\

3.  Category counts for categoricals\

4.  Box plots, histograms for vars of interest (Q-Q plots if you want)

```{r zion-vis, include=TRUE, fig.show='asis'}
# get descriptives for numerical vars
zion_tib %>% select(ss3.RAGtot,ss3.PTStot) %>% 
  psych::describe() %>% 
  knitr::kable(caption = "numerical variable descriptives", digits = 2) %>% 
  kableExtra::kable_classic(lightable_options = "hover")
# count categorical variables (example)
#zion_tib %>% count(ss1.Race)

# data visualization for relevant vars
p1 <- zion_tib %>% drop_na(ss3.RAGtot) %>%
  ggplot( aes(x=ss3.RAGtot)) + 
    geom_histogram(position = "identity", binwidth = .2, 
                   alpha = .5) + 
    theme_classic() + labs (title = "RAGtot histogram")
p2 <- zion_tib %>% #drop_na() %>%
    ggplot( aes(y=ss3.RAGtot)) + 
      geom_boxplot() + 
      theme_classic() +
      labs (title = "RAGtot box plot")
p1; p2

p3 <- zion_tib %>% drop_na(ss3.PTStot) %>%
  ggplot( aes(x=ss3.PTStot)) + 
    geom_histogram(position = "identity", binwidth = .2, 
                   alpha = .5) + 
    theme_classic() + labs (title = "RAGtot histogram")
p4 <- zion_tib %>% drop_na(ss3.PTStot) %>%
    ggplot( aes(y=ss3.PTStot)) + 
      geom_boxplot() + 
      theme_classic() +
      labs (title = "PTS box plot")
p3; p4


```

##### Statistical model

-   outcome: stress response (ss3.PTStot)\
-   explanatory vars (predictors): violence at t3 (ss3.RAGtot)

```{r zion-regression}

#statistical model
pts_lm <- zion_tib %>% drop_na(ss3.PTStot, ss3.RAGtot) %>% 
  lm(formula = ss3.PTStot ~ ss3.RAGtot)
summary(pts_lm)

#scatter plot with regression line - there are many points at x=0, so
#we will add a slight random jitter to the x values
p1z <- zion_tib %>% drop_na(ss3.PTStot, ss3.RAGtot) %>% 
  ggplot( aes(x=ss3.RAGtot, y=ss3.PTStot)) +
    geom_point(alpha = .25, position = position_jitter(width=.01)) +
    geom_smooth(method = "lm", formula = y ~ x) +
    theme_classic() +
    labs(title="stress by violent behavior - random jitter (.01) on x values")
p1z

#model diagnostics - residual plots
plot(pts_lm)

zion_tib %>% drop_na(ss3.PTStot, ss3.RAGtot) %>%
  select(ss3.PTStot, ss3.RAGtot) %>% 
  correlation::correlation(method = "pearson")
zion_tib %>% drop_na(ss3.PTStot, ss3.RAGtot) %>%
  select(ss3.PTStot, ss3.RAGtot) %>% 
  correlation::correlation(method = "kendall")

```

-   linear model - positive relation between violence and stress

    -   explore zeros in violent behavior

    -   for further analysis explore gender

------------------------------------------------------------------------

## Cory's data

```{r import-cory}

# import data
cory_tib <- readr::read_delim(
  "data/mccabecorym_344360_19967879_SRS_scores_og.txt",
  delim = "\t")
cory_tib <- cory_tib %>% 
  dplyr::mutate(
    ASD_fac = forcats::as_factor(ASD)
  ) %>% 
  dplyr::select(PID,ASD,ASD_fac,SRS_total:SRS_com)

```

#### Hypothesis/Research Question 1

-   what is the hypothesis examined? what columns are used to test the hypothesis

    -   Social impairment (SRS_total) is related ASD status

##### Descriptives and examine distributions of relevant variables

1.  Check cases and missing values, apply any exclusion criteria:

    -   Are missing values appearing appropriately? (scan the table after importing)\
    -   How many full cases?\
    -   Are you concerned about missing data?

2.  Mean, median, sd, min, max, cases for numeric variables of interest\

3.  Category counts for categoricals\

4.  Box plots, histograms for vars of interest (Q-Q plots if you want)

```{r cory-vis, include=TRUE, fig.show='asis'}
# get descriptives for numerical vars
cory_tib %>% select(SRS_total:SRS_com) %>% 
  psych::describe() %>% 
  knitr::kable(caption = "SRS descriptives", digits = 2) %>% 
  kableExtra::kable_classic(lightable_options = "hover")

# count categorical variables (example)
cory_tib %>% count(ASD_fac)

# data visualization for relevant vars
p1c <- cory_tib %>% #drop_na() %>%
  ggplot( aes(x=SRS_total, fill = ASD_fac)) + 
    geom_histogram(position = "identity", binwidth = 10, alpha = .25) + 
    theme_classic() + labs (title = "SRS_total histogram")
p1c
p2c <- cory_tib %>% #drop_na() %>%
    ggplot( aes(y=SRS_total, x = ASD_fac)) + geom_boxplot() + theme_classic() +
      labs (title = "SRS_total box plot")
p2c
p3c <- cory_tib %>% #drop_na() %>%
  ggplot( aes(x=SRS_cog, fill = ASD_fac)) + 
    geom_histogram(position = "identity", bins = 12, alpha = .25) + 
    theme_classic() + labs (title = "SRS_cog histogram")
p3c
p4c <- cory_tib %>% #drop_na() %>%
    ggplot( aes(y=SRS_cog, x = ASD_fac)) + geom_boxplot() + theme_classic() +
      labs (title = "SRS_cog box plot")
p4c
p5c <- cory_tib %>% #drop_na() %>%
  ggplot( aes(x=SRS_com, fill = ASD_fac)) + 
    geom_histogram(position = "identity", bins = 12, alpha = .25) + 
    theme_classic() + labs (title = "SRS_com histogram")
p5c
p6c <- cory_tib %>% #drop_na() %>%
    ggplot( aes(y=SRS_com, x = ASD_fac)) + geom_boxplot() + theme_classic() +
      labs (title = "SRS_com box plot")
p6c


```

##### Statistical model

-   mean SRS_total for ASD compared to no ASD
-   alternatively, logistic regression with ASD predicted by SRS_total

```{r cory-modeling}

# t-test of means for two groups:
cat("SRS_total t-test:\n")
cory_tib %>% t.test(formula= SRS_total ~ ASD_fac,
                    data = .,
                    paired = FALSE, 
                    alternative = "two.sided",
                    var.equal = FALSE
                    )
cat("\nSRS_cog t-test:\n")
cory_tib %>% t.test(formula= SRS_cog ~ ASD_fac,
                    data = .,
                    paired = FALSE, 
                    alternative = "two.sided",
                    var.equal = FALSE
                    )
cat("SRS_com t-test:\n")
cory_tib %>% t.test(formula= SRS_com ~ ASD_fac,
                    data = .,
                    paired = FALSE, 
                    alternative = "two.sided",
                    var.equal = FALSE
                    )

# effect sizes
cat("SRS_total cohens d:\n")
effectsize::cohens_d(SRS_total ~ ASD_fac,
                     data = cory_tib,
                     pooled_sd = TRUE,
                     paired = FALSE)
cat("\nSRS_cog cohens d:\n")
effectsize::cohens_d(SRS_cog ~ ASD_fac,
                     data = cory_tib,
                     pooled_sd = TRUE,
                     paired = FALSE)
cat("\nSRS_com cohens d:\n")
effectsize::cohens_d(SRS_com ~ ASD_fac,
                     data = cory_tib,
                     pooled_sd = TRUE,
                     paired = FALSE)
cat("\n")

```
