---
title: "Activity-Read data and examine distributions"
author: "JB"
date: "2/16/2021"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
library(tidyverse)
library(readr)
library(ggplot2)
```

#### 2.2 Read the ice bucket data and NHANES data into R

```{r Step2-load-datasets, results='hide', message=FALSE}
ice_tib <- readr::read_csv("data/ice_bucket.csv")
nhanes_tib <- readr::read_csv("data/nhanes_selectvars_n500.csv")
```

- 3.1.2 The warning message says "stat_bin()" using "bins = 30". Pick better value with "binwidth".

- 3.1.5 The distribution is unimodal, positively skewed

```{r Step3-icebucket-histogram3}
p1 <- ice_tib %>% 
      ggplot( aes(x=upload_day)) + geom_histogram() + theme_classic() +
        labs (title = "upload count by day (default bins)")
p1
p2 <- ice_tib %>% 
      ggplot( aes(x=upload_day)) + geom_histogram(binwidth = 1) + theme_classic() +
        labs (title = "upload count by day (1 bin per value)")
p2
p3 <- ice_tib %>% 
      ggplot( aes(x=upload_day)) +
        geom_histogram(binwidth = 10) + #or use breaks = seq(20,80,by= 10)
        theme_classic() +
        labs (title = "upload count by day (1 bin per 10 values)")
p3
```


4.1 The distribution is negatively skewed. The boxplot shows many outliers at low values because it includes a subset individuals with lower height (children)


```{r Step4.1-nhanes-height-all}
p1 <- nhanes_tib %>% drop_na(Height) %>%
      ggplot( aes(x=Height)) + geom_histogram(binwidth=2) + theme_classic() +
        labs (title = "Height distribution- all individuals")
p1
p2 <- nhanes_tib %>% drop_na(Height) %>%
      ggplot( aes(y=Height)) + geom_boxplot() + theme_classic()
p2
p3 <- nhanes_tib %>% drop_na(Height) %>%
      ggplot( aes(sample=Height)) + geom_qq() + geom_qq_line() + theme_classic()
p3
```


-   The distribution is roughly symmetrical, normal shaped

```{r Step4.2-nhanes-height-adults}
p1 <- nhanes_tib %>% drop_na(Height) %>% filter(Age>=18) %>%
      ggplot( aes(x=Height)) + geom_histogram(binwidth=2) + theme_classic() +
        labs (title = "Height distribution- Adults")
p1
p2 <- nhanes_tib %>% drop_na(Height) %>% filter(Age>=18) %>%
      ggplot( aes(y=Height)) + geom_boxplot() + theme_classic()
p2
p3 <- nhanes_tib %>% drop_na(Height) %>% filter(Age>=18) %>%
      ggplot( aes(sample=Height)) + geom_qq() + geom_qq_line() + theme_classic()
p3
```

#### 4.3 Mini-challenge

```{r Step4.3-nhanes-height-challenge}

p1 <- nhanes_tib  %>% drop_na(Height,Gender) %>% filter(Age>=18) %>%
      ggplot( aes(x=Height, fill=Gender)) + 
        geom_histogram(position="identity",alpha=.5,binwidth=2) + theme_classic() +
        labs (title = "Height distribution- Adults")
p1
p2 <- nhanes_tib %>% drop_na(Height,Gender) %>% filter(Age>=18) %>%
      ggplot( aes(y=Height, x=Gender)) + geom_boxplot() + theme_classic()
p2

```
